---
title: "descriptive analysis of solar panel data"
author: "Hannes Schoss"
output:
  pdf_document:
    toc: true
    number_sections: true
geometry: margin=1.5cm
header-includes:
  - \usepackage{booktabs}
  - \usepackage{float}
---

```{r setup, include=FALSE}
library(knitr)

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)

# setwd("/home/hannes/gh/solar-panels-r")
source("data-loader.R")

# helper functions for readable output
wh_to_mwh <- function(x) x / 1000000
wh_to_kwh <- function(x) x / 1000
pct <- function(x) round(100 * x, 1)
mwh <- function(x) round(wh_to_mwh(x), 3)
```

# introduction

Since 2022, my parents have had solar panels installed on their roof. This
document provides a descriptive analysis of the solar panel data and examines
several practical questions related to electricity production, household
consumption, grid usage, battery use, and seasonal variation.

A common assumption is that owning a solar panel system automatically leads to
substantial financial benefits. However, the actual value of such a system
depends on how much electricity is produced, how much of it is used directly,
how much is stored in the battery, and how much still has to be imported from
the public grid. This analysis therefore takes a data-based look at the
performance of the system and evaluates how efficiently the generated solar
energy is used.

The main goal is to better understand the cost-efficiency of the system. At the
same time, this can also be interpreted as a measure of ecological efficiency:
the more locally generated solar energy is used by the household, the less
dependent the household is on externally supplied electricity.

A second motivation is to assess the current degree of grid dependence. In
particular, the analysis investigates how much of the household’s electricity
demand can be covered by solar production and battery storage, and how much
still depends on electricity imported from the public grid. This is relevant
not only from a financial perspective, but also from a resilience perspective:
it helps answer whether the household could maintain its current electricity
use during a major disruption of the German energy infrastructure.


## structure of the data

Each row represents one aggregated 3-hour measurement interval. The timestamp
marks the beginning of the interval. Based on the hour of the timestamp, each
row is assigned to a daytime category.

```{r daytime}
dat$daytime <- NA

dat[dat$hour %in% c(21, 0, 3), ]$daytime <- "night"
dat[dat$hour == 6, ]$daytime <- "morning"
dat[dat$hour == 9, ]$daytime <- "late_morning"
dat[dat$hour == 12, ]$daytime <- "midday"
dat[dat$hour == 15, ]$daytime <- "afternoon"
dat[dat$hour == 18, ]$daytime <- "evening"
```

The most important variables are:

* `production` [Wh]: energy generated by the solar panels.
* `consumption` [Wh]: energy consumed by the household.
* `battery_charge` [Wh]: energy used to charge the 10 kWh battery.
* `battery_discharge` [Wh]: energy supplied by the battery to the household.
* `grid_feedin` [Wh]: energy exported to the public grid.
* `grid_consumption` [Wh]: energy imported from the public grid.
* `battery_state_of_charge` [%]: current battery charge level.
* `direct_consumption` [Wh]: solar energy consumed directly by the household.


## data cleaning

The raw dataset does not contain the same number of observations for every
year-month combination. This is expected in part, because months have different
numbers of days. As we already know a complete day contains eight observations.
With this information we can calculate that January should contain 31 × 8 = 248
observations, whereas February contains either 28 × 8 = 224 observations or, in
a leap year, 29 × 8 = 232 observations.

The following table shows the number of observations per year and month before
cleaning:

```{r raw-month-table}
table(dat$year, dat$month)
```

The table shows that some months are incomplete. March 2023 is incomplete
because data collection started during that month, so it is removed from the
analysis. 

```{r remove-incomplete-start-month}
dat <- dat[!(dat$year == 2023 & dat$month == 3), ]
```

```{r clean-data, echo=FALSE}
source("data-cleaner.R")
```

Other months contain only a small number of missing 3-hour intervals. For these
months, the missing values are approximated using the mean from the same month,
day, and hour in other years. In total, `r nrow(missing_slots)` out of 
`r nrow(full_grid)` are approximated.

```{r balanced-month-table}
table(dat$year, dat$month)
```

The resulting dataset preserves the correct calendar structure of the data while
removing distortions caused by incomplete months or missing 3-hour intervals.
This makes seasonal comparisons of absolute production and consumption values
more meaningful.



\newpage

# descriptive analysis

```{r monthly-energy-data, echo=FALSE}
flow_vars <- c(
  "production",
  "consumption",
  "battery_charge",
  "battery_discharge",
  "grid_feedin",
  "grid_consumption",
  "direct_consumption"
)

flow_labels <- c(
  production = "Solar production",
  consumption = "Total consumption",
  battery_charge = "Battery charge",
  battery_discharge = "Battery discharge",
  grid_feedin = "Grid feed-in",
  grid_consumption = "Grid consumption",
  direct_consumption = "Direct consumption"
)

totals_monthly <- aggregate(
  dat[flow_vars],
  by = list(
    year = dat$year,
    month = dat$month
  ),
  FUN = sum,
  na.rm = TRUE
)

mean_monthly <- aggregate(
  totals_monthly[flow_vars],
  by = list(month = totals_monthly$month),
  FUN = mean,
  na.rm = TRUE
)

season <- mean_monthly[
  c("month", "production", "grid_consumption")
]

n_years <- aggregate(
  year ~ month,
  data = totals_monthly,
  FUN = length
)

season <- merge(
  season,
  n_years,
  by = "month"
)

season$month_name <- month.name[season$month]

season$production_rank <- rank(
  -season$production,
  ties.method = "first"
)

season$grid_consumption_rank <- rank(
  -season$grid_consumption,
  ties.method = "first"
)

season$group <- ifelse(
  season$production_rank <= 6,
  "warm",
  "cold"
)

season <- season[order(season$production_rank), ]

totals_monthly$group <- season$group[
  match(totals_monthly$month, season$month)
]
```

First I want to look at the data from a yearly perspective, and decide
if one can find categories like "seasons" in the data.
Because months have different numbers of days, I use average daily
values for monthly comparisons instead of simple monthly sums.

```{r monthly-plots, fig.width=9, fig.height=7, echo = FALSE}
par(mfrow = c(2, 2))

source("functions/plot-month-by-year.R")

plot_month_by_year(
  dat, "production",
  ylab = "Average daily production [kWh/day]",
  main = "Solar Production"
)

plot_month_by_year(
  dat, "consumption",
  ylab = "Average daily consumption [kWh/day]",
  main = "Household Consumption"
)

plot_month_by_year(
  dat, "grid_feedin",
  ylab = "Average daily feed-in [kWh/day]",
  main = "Grid Feed-in"
)

plot_month_by_year(
  dat, "grid_consumption",
  ylab = "Average daily grid consumption [kWh/day]",
  main = "Grid Consumption"
)

par(mfrow = c(1, 1))
```

The plots show strong seasonality in the energy flows. Solar production and
grid feed-in peak in the brighter months and decline sharply in winter. As a
result, grid consumption remains very low during spring and summer, when solar
availability is high, and rises again in autumn and winter as solar production
declines.

Grid consumption and Solar Production / Grid Feed-in seem to be inverse. The
Household Consumption stays on a consistent level all year round. One notable
exception is the pronounced spike in household consumption in the 2025 and 2026
data. Although this pattern appears unusual at first glance, it has a specific
explanation that will be addressed in the financial section of this paper.

## defining solar seasons

Instead of using fixed calendar seasons such as the astronomical summer begin,
I define seasons directly from the data as high-production and low-production
solar months.

* **warm months**: the six months with the highest total solar production
* **cold months**: the six months with the lowest total solar production

```{r solar-seasons, echo=FALSE, results='asis'}
knitr::kable(
  season,
  format = "latex",
  booktabs = TRUE,
  digits = 0,
  row.names = FALSE,
  caption = paste(
    "Estimated average monthly solar production and grid consumption,",
    "ranked by average monthly production"
  )
)
```

```{r cor-prod-grid-con, echo = FALSE}
cor_prod_grid_con <- cor(
  season$production_rank,
  season$grid_consumption_rank,
  method = "spearman"
)

```

The table ranks months by total solar production. Months with ranks 1 to 6 are
classified as warm months, while the remaining months are classified as cold
months. To evaluate whether high-production months also tend to have lower grid
consumption, I calculated the correlation between the production ranking and
the grid consumption ranking, which is known as the Spearman Correlation. The
resulting correlation is `r round(cor_prod_grid_con, 2)` and very high.

# household energy balance

This section digs into the total energy flows over the full observation period.
First we look at the the absolute energy amounts in MWh. Then we look at the 
energy flow, comparing the cold and the warm months.


## total energy balance

The total values provide an overview of the household energy balance, but they
hide important seasonal differences. Solar production is expected to be higher
during the warm season, while grid consumption is usually more relevant during
the cold season. Because of this, the energy flows were summarized separately for
warm and cold months.

```{r total-data, echo=FALSE, results='asis', fig.pos = 'H'}
total <- data.frame(
  energy_flow = unname(flow_labels),
  mwh = as.numeric(
    wh_to_mwh(colSums(dat[flow_vars], na.rm = TRUE))
  )
)

knitr::kable(
  total,
  format = "latex",
  booktabs = TRUE,
  digits = 2,
  row.names = FALSE,
  col.names = c("Energy flow", "MWh"),
  caption = "Total energy flows over the full observation period."
)
```

## seasonal breakdown of household energy flows

```{r seasonal-flowchart,
  echo=FALSE,
  fig.width=9,
  fig.height=5,
  fig.pos='H',
  fig.cap="Average monthly energy flows in warm and cold months."
}
source("functions/plot-flowchart.R")
source("functions/plot-flowchart.R")

mean_seasonal_month <- aggregate(
  totals_monthly[flow_vars],
  by = list(group = totals_monthly$group),
  FUN = mean,
  na.rm = TRUE
)

mean_seasonal_month[flow_vars] <- lapply(
  mean_seasonal_month[flow_vars],
  wh_to_mwh
)

warm_flow <- mean_seasonal_month[
  mean_seasonal_month$group == "warm",
  flow_vars,
  drop = FALSE
]

cold_flow <- mean_seasonal_month[
  mean_seasonal_month$group == "cold",
  flow_vars,
  drop = FALSE
]

plot_flowchart(
  warm_flow,
  "Average monthly energy flows in high-production months"
)

plot_flowchart(
  cold_flow,
  "Average monthly energy flows in low-production months"
)
```





```{r berechnung}
# flow_vars <- colnames(total)

flow_vars <- c(
  "production",
  "consumption",
  "battery_charge",
  "battery_discharge",
  "grid_feedin",
  "grid_consumption",
  "direct_consumption"
)

totals_monthly <- aggregate(
  dat[flow_vars], by = list(year = dat$year, month = dat$month),
  FUN = sum, na.rm = TRUE)

totals_monthly$season <- ifelse(
  totals_monthly$month %in% 4:9,
  "warm",
  "cold"
)

mean_by_month <- aggregate(
  totals_monthly[flow_vars],
  by = list(
    month = totals_monthly$month,
    season = totals_monthly$season
  ),
  FUN = mean,
  na.rm = TRUE
)

mean_by_season <- aggregate(
  mean_by_month[flow_vars],
  by = list(season = mean_by_month$season),
  FUN = mean,
  na.rm = TRUE
)
```


```{r total-flowchart, echo=FALSE, fig.width=9, fig.height=5, fig.pos='H', fig.cap="Total energy flows in warm and cold months."}
warm_monthly_mean <- mean_by_season[
  mean_by_season$season == "warm",
  flow_vars
]

cold_monthly_mean <- mean_by_season[
  mean_by_season$season == "cold",
  flow_vars
]

warm_monthly_mean[] <- lapply(warm_monthly_mean, wh_to_mwh)
cold_monthly_mean[] <- lapply(cold_monthly_mean, wh_to_mwh)

plot_flowchart(
  warm_monthly_mean,
  "Average monthly energy flows in warm months"
)

plot_flowchart(
  cold_monthly_mean,
  "Average monthly energy flows in cold months"
)
```





## financial implications of seasonality

As one might guess from the visualizations, there seem to be different 
money gains from the tow  different seasons in the year.

lets see how much the system earned and spent every year for the cold and the warm season.

earning: 5 cent per kwh
spending: 35 cent per kwh


ist es ausgeglichen 



Here, the net balance is defined as total solar production minus total household consumption. A positive value means that the system produced more energy than the household consumed over the full observation period.
what was the money balance?


The financial implications of grid exchange are strongly asymmetric because
electricity bought from the grid is much more expensive than electricity sold
to the grid. Assuming a feed-in compensation of 0.05 EUR/kWh and a grid
electricity price of 0.35 EUR/kWh, one kilowatt-hour imported from the grid
costs seven times as much as one kilowatt-hour exported to the grid earns.

In the warm months, grid feed-in generates about 395 EUR of revenue, while grid
consumption costs only about 48 EUR. This results in a positive grid-related
balance of about 347 EUR. In the cold months, however, grid feed-in generates
only about 59 EUR, while grid consumption costs about 256 EUR, resulting in a
negative balance of about 197 EUR.

This shows why direct consumption and battery use are financially more valuable
than feeding electricity into the grid. Each kilowatt-hour used directly by the
household avoids buying electricity for 0.35 EUR, whereas feeding the same
kilowatt-hour into the grid would only earn 0.05 EUR. Therefore, using solar
electricity locally is worth about 0.30 EUR more per kWh than exporting it.



total_warm$grid_feedin * 1000 * 0.05

total_cold$grid_feedin * 1000 * 0.05

total_warm$grid_consumption * 1000 * 0.35
total_cold$grid_consumption * 1000 * 0.35


if feeding in and drawing were balanced
total_warm$grid_feedin * 1000 * 0.35



```{r -net-balance}
net_energy_balance_mwh <- wh_to_mwh(
  sum(dat$production, na.rm = TRUE) - sum(dat$consumption, na.rm = TRUE)
)

net_grid_balance_mwh <- wh_to_mwh(
  sum(dat$grid_feedin, na.rm = TRUE) - sum(dat$grid_consumption, na.rm = TRUE)
)

data.frame(
  balance = c("Production minus consumption", "Grid export minus grid import"),
  energy_MWh = c(net_energy_balance_mwh, net_grid_balance_mwh)
)
```

**Answer:** The production-minus-consumption balance is about `r net_energy_balance_mwh` MWh. The grid-export-minus-grid-import balance is about `r net_grid_balance_mwh` MWh. This means the household produced substantially more solar energy than it consumed, but part of this surplus was exported because production and demand did not always occur at the same time.

## 7. What percentage of total solar production was used locally?

```{r -local-use}
local_solar_use_wh <- sum(dat$production, na.rm = TRUE) - sum(dat$grid_feedin, na.rm = TRUE)
local_solar_share <- local_solar_use_wh / sum(dat$production, na.rm = TRUE)
local_solar_share_pct <- pct(local_solar_share)
local_solar_share_pct
```

