---
title: "descriptive analysis of solar panel data"
author: "Hannes Schoss"
output:
  pdf_document:
    toc: true
    number_sections: true
geometry: margin=1.5cm
header-includes:
  - \usepackage{booktabs}
  - \usepackage{float}
  - \usepackage{placeins}
---

```{r setup, include=FALSE}
library(knitr)

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.pos = "H"
)

# setwd("/home/hannes/gh/solar-panels-r")
source("data-loader.R")

wh_to_mwh <- function(x) x / 1000000
wh_to_kwh <- function(x) x / 1000
pct <- function(x) round(100 * x, 1)

flow_vars <- c(
  "production",
  "consumption",
  "battery_charge",
  "battery_discharge",
  "grid_feedin",
  "grid_consumption",
  "direct_consumption"
)

flow_labels <- c(
  production = "Solar production",
  consumption = "Total consumption",
  battery_charge = "Battery charge",
  battery_discharge = "Battery discharge",
  grid_feedin = "Grid feed-in",
  grid_consumption = "Grid consumption",
  direct_consumption = "Direct consumption"
)

energy_table <- function(values_wh) {
  data.frame(
    energy_flow = unname(flow_labels),
    mwh = as.numeric(wh_to_mwh(values_wh[flow_vars])),
    row.names = NULL
  )
}

convert_flows_to_mwh <- function(data) {
  data[flow_vars] <- lapply(data[flow_vars], wh_to_mwh)
  data
}
```

# introduction

Since 2022, my parents have had solar panels installed on their roof. This
report provides a descriptive analysis of the solar panel data and examines
several practical questions related to electricity production, household
consumption, grid usage, battery use, and seasonal variation.

A common assumption is that owning a solar panel system automatically leads to
substantial financial benefits. However, the actual value of such a system
depends on how much electricity is produced, how much of it is used directly,
how much is stored in the battery, and how much still has to be imported from
the public grid. This analysis therefore takes a data-based look at the system
and evaluates how efficiently the generated solar energy is used.

The main goal is to better understand the cost-efficiency of the system. At the
same time, this can also be interpreted as a measure of ecological efficiency:
the more locally generated solar energy is used by the household, the less
dependent the household is on externally supplied electricity.

A second motivation is to assess the current degree of grid dependence. In
particular, the analysis investigates how much of the household's electricity
demand can be covered by solar production and battery storage, and how much
still depends on electricity imported from the public grid.

## structure of the data

Each row represents one aggregated 3-hour measurement interval. The timestamp
marks the beginning of the interval. Based on the hour of the timestamp, each
row is assigned to a daytime category.

```{r daytime}
dat$daytime <- NA

dat[dat$hour %in% c(21, 0, 3), ]$daytime <- "night"
dat[dat$hour == 6, ]$daytime <- "morning"
dat[dat$hour == 9, ]$daytime <- "late_morning"
dat[dat$hour == 12, ]$daytime <- "midday"
dat[dat$hour == 15, ]$daytime <- "afternoon"
dat[dat$hour == 18, ]$daytime <- "evening"
```

The most important variables are:

* `production` [Wh]: energy generated by the solar panels.
* `consumption` [Wh]: energy consumed by the household.
* `battery_charge` [Wh]: energy used to charge the 10 kWh battery.
* `battery_discharge` [Wh]: energy supplied by the battery to the household.
* `grid_feedin` [Wh]: energy exported to the public grid.
* `grid_consumption` [Wh]: energy imported from the public grid.
* `battery_state_of_charge` [%]: current battery charge level.
* `direct_consumption` [Wh]: solar energy consumed directly by the household.

## data cleaning

The raw dataset does not contain the same number of observations for every
year-month combination. This is expected in part, because months have different
numbers of days. A complete day contains eight observations, so January should
contain 31 * 8 = 248 observations, whereas February contains either 28 * 8 =
224 observations or, in a leap year, 29 * 8 = 232 observations.

The following table shows the number of observations per year and month before
cleaning:

```{r raw-month-table}
table(dat$year, dat$month)
```

The table shows that some months are incomplete. March 2023 is incomplete
because data collection started during that month, so it is removed from the
analysis.

```{r remove-incomplete-start-month}
dat <- dat[!(dat$year == 2023 & dat$month == 3), ]
```

```{r clean-data, echo=FALSE}
source("data-cleaner.R")
```

Other months contain only a small number of missing 3-hour intervals. For these
months, the missing values are approximated using the mean from the same month,
day, and hour in other years. In total, `r nrow(missing_slots)` out of
`r nrow(full_grid)` are approximated.

```{r balanced-month-table}
table(dat$year, dat$month)
```

The resulting dataset preserves the correct calendar structure of the data while
removing distortions caused by incomplete months or missing 3-hour intervals.
This makes seasonal comparisons of absolute production and consumption values
more meaningful.

\newpage

# descriptive analysis

```{r energy-setup, echo=FALSE}
totals_monthly <- aggregate(
  dat[flow_vars],
  by = list(
    year = dat$year,
    month = dat$month
  ),
  FUN = sum,
  na.rm = TRUE
)

mean_monthly <- aggregate(
  totals_monthly[flow_vars],
  by = list(month = totals_monthly$month),
  FUN = mean,
  na.rm = TRUE
)

season <- mean_monthly[c("month", "production", "grid_consumption")]

n_years <- aggregate(
  year ~ month,
  data = totals_monthly,
  FUN = length
)

names(n_years)[names(n_years) == "year"] <- "n_years_observed"

season <- merge(
  season,
  n_years,
  by = "month"
)

season$month_name <- month.name[season$month]

season$production_rank <- rank(
  -season$production,
  ties.method = "first"
)

season$grid_consumption_rank <- rank(
  -season$grid_consumption,
  ties.method = "first"
)

season$group <- ifelse(
  season$production_rank <= 6,
  "warm",
  "cold"
)

season <- season[order(season$production_rank), ]

totals_monthly$group <- season$group[
  match(totals_monthly$month, season$month)
]
```

First, the data are inspected from a yearly perspective to identify whether
seasonal categories can be derived from the data. Because months have different
numbers of days, the monthly plots use average daily values instead of simple
monthly sums.

```{r monthly-plots, fig.width=9, fig.height=7, echo=FALSE}
par(mfrow = c(2, 2))

source("functions/plot-month-by-year.R")

plot_month_by_year(
  dat,
  "production",
  ylab = "Average daily production [kWh/day]",
  main = "Solar production"
)

plot_month_by_year(
  dat,
  "consumption",
  ylab = "Average daily consumption [kWh/day]",
  main = "Household consumption"
)

plot_month_by_year(
  dat,
  "grid_feedin",
  ylab = "Average daily feed-in [kWh/day]",
  main = "Grid feed-in"
)

plot_month_by_year(
  dat,
  "grid_consumption",
  ylab = "Average daily grid consumption [kWh/day]",
  main = "Grid consumption"
)

par(mfrow = c(1, 1))
```

The plots show strong seasonality in the energy flows. Solar production and
grid feed-in peak in the brighter months and decline sharply in winter. As a
result, grid consumption remains very low during spring and summer, when solar
availability is high, and rises again in autumn and winter as solar production
declines.

Grid consumption and solar production appear to move in opposite directions,
while household consumption remains comparatively stable over the year. One
notable exception is the pronounced spike in household consumption in the 2025
and 2026 data. This pattern is addressed in the financial section of the report.

## Defining solar seasons

Instead of using fixed calendar seasons, seasons are defined directly from the
data as high-production and low-production solar months.

* **Warm months**: the six months with the highest average monthly solar
  production.
* **Cold months**: the six months with the lowest average monthly solar
  production.

The average monthly values are calculated in two steps. First, production and
grid consumption are summed within each year-month combination. Second, these
monthly totals are averaged across years. This avoids giving more weight to
years with more observations.

```{r solar-seasons, echo=FALSE, results='asis'}
knitr::kable(
  season,
  format = "latex",
  booktabs = TRUE,
  digits = 0,
  row.names = FALSE,
  caption = paste(
    "Estimated average monthly solar production and grid consumption,",
    "ranked by average monthly production"
  )
)
```

```{r cor-prod-grid-con, echo=FALSE}
cor_prod_grid_con <- cor(
  season$production_rank,
  season$grid_consumption_rank,
  method = "spearman"
)
```

The table ranks months by average monthly solar production. Months with ranks
1 to 6 are classified as warm months, while the remaining months are classified
as cold months. To evaluate whether high-production months also tend to have
lower grid consumption, I calculated the Spearman correlation between the
production ranking and the grid-consumption ranking. The resulting correlation
is `r round(cor_prod_grid_con, 2)`, which indicates a strong inverse seasonal
relationship between solar production and grid dependence.

# household energy balance

This section summarizes the household energy balance. First, it presents total
energy flows over the full observation period. Then, it shows how these flows
differ between warm and cold solar seasons.

## total energy balance

The total values provide an overview of the full household energy balance.
However, they do not show how much of the total is driven by warm or cold
months. For that reason, the total table is followed by a seasonal breakdown.

```{r total-data, echo=FALSE, results='asis'}
total_wh <- colSums(dat[flow_vars], na.rm = TRUE)
total <- energy_table(total_wh)

knitr::kable(
  total,
  format = "latex",
  booktabs = TRUE,
  digits = 2,
  row.names = FALSE,
  col.names = c("Energy flow", "MWh"),
  caption = "Total energy flows over the full observation period."
)
```

## seasonal breakdown of household energy flows

The total values hide important seasonal differences. Solar production is
expected to be higher during warm months, while grid consumption is usually
more relevant during cold months.

To compare seasonal energy flows without bias from different observation
lengths, the analysis first aggregates all 3-hour observations into monthly
totals. These monthly totals are then grouped into warm and cold seasons for
each year. Incomplete seasons are standardized to a six-month equivalent before
the seasonal totals are averaged across years. The flowcharts therefore show
the average total energy flows of a typical warm or cold season, not average
monthly values and not simple totals over the observed period.

```{r seasonal-flow-data, echo=FALSE}
seasonal_totals <- aggregate(
  totals_monthly[flow_vars],
  by = list(
    year = totals_monthly$year,
    group = totals_monthly$group
  ),
  FUN = sum,
  na.rm = TRUE
)

season_months <- aggregate(
  month ~ year + group,
  data = totals_monthly,
  FUN = length
)

names(season_months)[names(season_months) == "month"] <- "n_months"

seasonal_totals <- merge(
  seasonal_totals,
  season_months,
  by = c("year", "group")
)

seasonal_totals[flow_vars] <- seasonal_totals[flow_vars] /
  seasonal_totals$n_months * 6

mean_seasonal_total <- aggregate(
  seasonal_totals[flow_vars],
  by = list(group = seasonal_totals$group),
  FUN = mean,
  na.rm = TRUE
)

mean_seasonal_total <- convert_flows_to_mwh(mean_seasonal_total)

warm_flow <- mean_seasonal_total[
  mean_seasonal_total$group == "warm",
  flow_vars,
  drop = FALSE
]

cold_flow <- mean_seasonal_total[
  mean_seasonal_total$group == "cold",
  flow_vars,
  drop = FALSE
]
```

```{r warm-flowchart, echo=FALSE, fig.width=6, fig.height=4, out.width="0.82\\linewidth", fig.pos="H", fig.align="center", fig.cap="Average six-month energy flows in the warm season."}
source("functions/plot-flowchart.R")

plot_flowchart(
  warm_flow,
  "Warm season"
)
```

```{r cold-flowchart, echo=FALSE, fig.width=6, fig.height=4, out.width="0.82\\linewidth", fig.pos="H", fig.align="center", fig.cap="Average six-month energy flows in the low-production season."}
plot_flowchart(
  cold_flow,
  "Low-production season"
)
```

\FloatBarrier


# financial implications of seasonality

The financial implications of grid exchange are asymmetric because electricity
bought from the grid is much more expensive than electricity sold to the grid.
Assuming a feed-in compensation of 0.05 EUR/kWh and a grid electricity price of
0.35 EUR/kWh, one kilowatt-hour imported from the grid costs seven times as
much as one kilowatt-hour exported to the grid earns.

```{r finance-by-season, echo=FALSE, results='asis', fig.pos='H'}
feedin_rate <- 0.05
grid_rate <- 0.35

finance_by_season <- data.frame(
  season = mean_seasonal_total$group,
  feedin_revenue_eur = mean_seasonal_total$grid_feedin * 1000 * feedin_rate,
  grid_cost_eur = mean_seasonal_total$grid_consumption * 1000 * grid_rate
)

finance_by_season$net_grid_balance_eur <-
  finance_by_season$feedin_revenue_eur - finance_by_season$grid_cost_eur

knitr::kable(
  finance_by_season,
  format = "latex",
  booktabs = TRUE,
  digits = 2,
  row.names = FALSE,
  col.names = c(
    "Season",
    "Feed-in revenue [EUR]",
    "Grid cost [EUR]",
    "Net grid balance [EUR]"
  ),
  caption = paste(
    "Estimated grid-related financial balance for an average",
    "six-month warm or cold season."
  )
)
```

This shows why direct consumption and battery use are financially more valuable
than feeding electricity into the grid. Each kilowatt-hour used directly by the
household avoids buying electricity for 0.35 EUR, whereas feeding the same
kilowatt-hour into the grid would only earn 0.05 EUR. Therefore, using solar
electricity locally is worth about 0.30 EUR more per kWh than exporting it.

## Net energy balances

Here, the net production balance is defined as total solar production minus
total household consumption. A positive value means that the system produced
more energy than the household consumed over the full observation period.

```{r net-balance, echo=FALSE, results='asis'}
net_energy_balance_mwh <- wh_to_mwh(
  sum(dat$production, na.rm = TRUE) -
    sum(dat$consumption, na.rm = TRUE)
)

net_grid_balance_mwh <- wh_to_mwh(
  sum(dat$grid_feedin, na.rm = TRUE) -
    sum(dat$grid_consumption, na.rm = TRUE)
)

net_balance <- data.frame(
  balance = c(
    "Production minus consumption",
    "Grid export minus grid import"
  ),
  energy_mwh = c(
    net_energy_balance_mwh,
    net_grid_balance_mwh
  )
)

knitr::kable(
  net_balance,
  format = "latex",
  booktabs = TRUE,
  digits = 2,
  row.names = FALSE,
  col.names = c("Balance", "MWh"),
  caption = "Net energy balances over the full observation period."
)
```

The production-minus-consumption balance is about
`r round(net_energy_balance_mwh, 2)` MWh. The grid-export-minus-grid-import
balance is about `r round(net_grid_balance_mwh, 2)` MWh. This means the
household produced substantially more solar energy than it consumed, but part
of this surplus was exported because production and demand did not always occur
at the same time.

## Local use of solar production

```{r local-use, echo=FALSE, results='asis'}
local_solar_use_wh <- sum(dat$production, na.rm = TRUE) -
  sum(dat$grid_feedin, na.rm = TRUE)

local_solar_share <- local_solar_use_wh /
  sum(dat$production, na.rm = TRUE)

local_use <- data.frame(
  measure = "Share of solar production used locally",
  percent = pct(local_solar_share)
)

knitr::kable(
  local_use,
  format = "latex",
  booktabs = TRUE,
  digits = 1,
  row.names = FALSE,
  col.names = c("Measure", "Percent"),
  caption = "Share of solar production used locally by the household."
)
```

The household used about `r pct(local_solar_share)` percent of solar production
locally, either directly or through the battery. The remaining share was fed
into the public grid.
